Analysis-ready VCF at Biobank scale using Zarr

### 2025-01-31 paper.tex
Correct calcs and typo

% >>> 78_195 * 722*10**6 * 2 / 1024**4
% 102.69430276821367
% >>> (78_195 * 722*10**6 * 2 / 1024**3) / 25.5
% 4123.881020966698
% >>> 4123.881020966698 / 360
% 1.1455
% >>> 4123.881020966698 / 60
% 68.73135034944497
% >>> 1.1455 * 4.848
% 5.553383999999999
Returning to Fig~\ref{fig-s3-throughput}, we can see that the af-dist
calculation is performed at a maximum rate of 25.5 GiB/s using 128
processes. Genome wide, the aggV2 dataset consists of
722 million variants for 78,195 samples, giving a total of around
103 TiB of genotype data. We could therefore process the entire
dataset at this maximum rate in about 69 minutes, with a compute
cost of \$5.55
(on-demand hourly rate of \$4.848 for a c5.24xlarge instance
in the eu-west-2 region, as-of 2025-01-31~\cite{ec2-pricing}).
AWS and other major clouds charge per object request (i.e. without
considering the number of bytes transferred) and so the IO
cost of this operation is determined by the number of chunks.
% >>> (10_000 * 1000 * 2) / 1024**2
% 19.073486328125
% >>> (78_195 * 722*10**6 * 2) / (10_000 * 1000 * 2)
% 5645679.0
% >>> (5645679.0 / 1000) *0.00042
% 2.3711851800000003
As each 10,000 $\times$ 1,000 chunk encodes 19 MiB of 
genotype data, we incur about 5.6 million object requests
to retrieve all chunks. 
At \$0.00042 per 1000 GET requests for S3 Standard
storage 
(eu-west-2 region as-of 2025-01-31~\cite{s3-pricing}), 
this comes to \$2.37.
Thus, it should be possible to perform af-dist and similar
calculations genome wide on aggV2 for less than \$10.
Assuming 50KiB per chunk (the average stored chunk size 
for chromosome 2 is 47.57 KiB) we would need a total
of 269 GiB to store the genotype data for the entire dataset.
% >>> (5645679.0 * 50 * 1024) / 1024**3
% 269.20695304870605
% 269.20695304870605 * 0.024 * 12
% 77.53160247802734
At \$0.024 per GiB per month for S3 Standard
(eu-west-2 region as-of 2025-01-31~\cite{s3-pricing})
this comes to a yearly storage cost of \$77.53
(note that this is just the genotype data, not including
other, much larger, fields 
listed in Table~\ref{tab-genomics-england-data}).

One of our benchmarks in the Genomics England case
study section was to compare to compare a straightforward
sequential Zarr implementation of af-dist with
\texttt{bcftools} 
on the genome slice corresponding to one of the original
106 VCF files for chromosome 2. There, Zarr's efficient data format
allows us to perform the same calculation in 4.25
% >>> 716.3 / 60
% 11.938333333333333
minutes vs almost 12 hours for VCF. 
Here, by adapting our code to the 
high-throughput, high-variance nature of cloud stores and 
using the natural chunk-based parallelism of Zarr
we reduce the elapsed time to 4.15 seconds.
These benchmarks illustrate the transformative 
potential of Zarr and cloud computing.

>>>>

% >>> 78_195 * 722*10**6 * 2 / 1024**4
% 102.69430276821367
% >>> (78_195 * 722*10**6 * 2 / 1024**3) / 25.5
% 4123.881020966698
% >>> 4123.881020966698 / 360
% 1.1455
% >>> 4123.881020966698 / 60
% 68.73135034944497
% >>> 1.1455 * 4.848
% 5.553383999999999
Returning to Fig~\ref{fig-s3-throughput}, we can see that the af-dist
calculation is performed at a maximum rate of 25.5 GiB/s using 128
processes. Genome wide, the aggV2 dataset consists of
722 million variants for 78,195 samples, giving a total of around
103 TiB of genotype data. We could therefore process the entire
dataset at this maximum rate in about 69 minutes, with a compute
cost of \$5.55
(on-demand hourly rate of \$4.848 for a c5.24xlarge instance
in the eu-west-2 region, as-of 2025-01-31~\cite{ec2-pricing}).
AWS and other major clouds charge per object request (i.e. without
considering the number of bytes transferred) and so the IO
cost of this operation is determined by the number of chunks.
% >>> (10_000 * 1000 * 2) / 1024**2
% 19.073486328125
% ceil here as a whole chunk is always used
% >>> ceil(78_195/1_000) * ceil(722*10**6/10_000)
% 5703800
% >>> (5703800 / 1000) *0.00042
% 2.3955960000000003
As each 10,000 $\times$ 1,000 chunk encodes 19 MiB of 
genotype data, we incur about 5.7 million object requests
to retrieve all chunks. 
At \$0.00042 per 1000 GET requests for S3 Standard
storage 
(eu-west-2 region as-of 2025-01-31~\cite{s3-pricing}), 
this comes to \$2.39.
Thus, it should be possible to perform af-dist and similar
calculations genome wide on aggV2 for less than \$10.
Assuming 50KiB per chunk (the average stored chunk size 
for chromosome 2 is 47.57 KiB) we would need a total
of 272 GiB to store the genotype data for the entire dataset.
% the boundary chunks where a whole chunk is not used will be smaller
% but this is a tiny effect.
% >>> (5703800.0 * 50 * 1024) / 1024**3
% 271.97837829589844
% 271.97837829589844 * 0.024 * 12
% 78.32977294921875
At \$0.024 per GiB per month for S3 Standard
(eu-west-2 region as-of 2025-01-31~\cite{s3-pricing})
this comes to a yearly storage cost of \$78.33
(note that this is just the genotype data, not including
other, much larger, fields 
listed in Table~\ref{tab-genomics-england-data}).

One of our benchmarks in the Genomics England case
study section was to compare a straightforward
sequential Zarr implementation of af-dist with
\texttt{bcftools} 
on the genome slice corresponding to one of the original
106 VCF files for chromosome 2. There, Zarr's efficient data format
allows us to perform the same calculation in 4.25
% >>> 716.3 / 60
% 11.938333333333333
minutes vs almost 12 hours for VCF. 
Here, by adapting our code to the 
high-throughput, high-variance nature of cloud stores and 
using the natural chunk-based parallelism of Zarr
we reduce the elapsed time to 4.15 seconds.
These benchmarks illustrate the transformative 
potential of Zarr and cloud computing.
